Gaia's constraints on the very massive $\sim$$150\,M_\odot$ runaway star candidate VFTS682

### 2018-07-26 VFTS682_APJL.tex
added ref. to drew+18
started working on Selma's comments

Figure~\ref{fig:dist} illustrates the kinematic properties of VFTS OB-type
and Wolf-Rayet stars in the 30 Doradus region with \emph{Gaia} DR2 errors on the proper motion
components of less than $0.1\,\mathrm{mas\ yr^{-1}}$ (dark blue lines,
including VFTS682), and less than $0.05\,\mathrm{mas\ yr^{-1}}$ (light hatched
blue). The green vertical bands show the expectations for VFTS682 if
it were ejected from R136 early in its lifetime. We highlight this star together with the two O-type
runaways, VFTS16 and VFTS72, \mbox{(re-)identified} by \cite{lennon:18}.

The left panel shows the distribution in proper motion relative
to R136 from \emph{Gaia} DR2, selecting stars with small error bars only. Although VFTS682 is not peculiar, its relative proper motion
falls in the range expected for it to reach its present day isolation
within its apparent age (green vertical band). This suggests it is indeed a ``slow-runaway'' as
suggested by \cite{bestenlehner:11}. Subtracting
the mean motion of R136, we obtain relative proper motions (projected
velocities) of $\delta \mu_{Gaia}=0.13\pm 0.09\masyr (32\pm 21\kms)$ and
$\delta \mu_{HST}=0.20\pm 0.10\masyr (47 \pm 24\kms )$. Both values
are consistent with each other, but also with no relative motion
within $2\sigma$.

@@

The right panel shows the distributions in projected flight direction for these
stars: $\theta$ is the angle between the projected radial direction from the
center of R136 to each star and its projected relative proper motion (see also \Figref{fig:main}). Dynamical ejections from the
cluster should %are expected to
produce
close to radial ejections, i.e.~$\theta\simeq0$. On this panel, the green band highlights
% the range of angles corresponding to
an opening angle around the
projected radial direction of 45\,degrees. The direction of motion of VFTS682 % is
% poorly constrained because of the large errors on the proper motion
% components both in HST and \emph{Gaia} data, but
% it
is consistent with dynamical ejection from R136. We note that
VFTS72 has a small radial velocity, while VFTS16 (and possibly VFTS682) has
large peculiar radial velocity and therefore accurate distances along
the line of sight are
needed to contrain the flight direction in three dimensions.

@@

Therefore, although the motion of VFTS682 is not an outlier compared
to other massive stars with small proper motion errors in \emph{Gaia} DR2, it is possible to tentatively claim that it is
a slow runaway with velocity compatible with dynamical ejection from
R136 early in the evolution of both the star and the cluster.

\begin{figure}%[tbp]
  \centering
  \includegraphics[width=0.48\textwidth]{./figures/Figure2_paper_vfts682_hst_corr_nor136err}  
  \caption{The thick yellow arrows indicate proper motions relative R136 (blue
    circle) from \emph{Gaia} DR2, multiplied by 0.4\,Myr. The prolongations in the opposite
    direction are proportional to the apparent ages of the stars, and
    the thin lines illustrate the error cone on the potential
    origin. The green thick arrow
    indicates the HST proper motion of VFTS682, with the relative
    error cone and prolongation proportional to its apparent
    age.}

  \label{fig:main}
\end{figure}

@@

In \citet{lennon:18} we carried out a similar study on the fast
moving O-type stars in the region, and found two very massive runaway stars in the 30 Doradus region. One of them (VFTS 16)
was previously known as a runaway star from its line of sight velocity
\citep[][]{evans:10}. \citet{lennon:18} also concluded that VFTS16 is 
the result of a dynamical ejection from the R136 cluster, while the
origin of the other star (VFTS72) is less clear given its direction
of motion (cf.~\Figref{fig:dist}). The value of $\tau_\mathrm{kin}\simeq0.9$\,Myr we find for
VFTS682 is smaller than the
corresponding value for VFTS16 of $\sim$1.5\,Myr, possibly in tension with the apparent age of that star. This means that the more
massive VFTS682 was ejected later than VFTS16 from the same cluster.

>>>>

The black points in~\Figref{fig:dist} show the proper motion 
(left panel) and the projected flight direction (right panel) relative
to R136 of
VFTS682. $\theta$ is the angle between the projected radial direction from the
center of R136 to each star and its projected relative proper motion
(see also \Figref{fig:main}). Dynamical ejections from the
cluster should %are expected to
produce
close to radial ejections, i.e.~$\theta\simeq0$. The green vertical bands highlight the expectations for these
two quantities if VFTS682 were ejected from R136 early in its
lifetime. Although the error bars are substantial and VFTS682 is not
an outlier, the agreement
suggests that the star is indeed a ``slow-runaway'' as suggested by
\cite{bestenlehner:11}. For comparison, we also show the relative
proper motion of VFTS16 and VFTS72 (gray points), and the distribution
in relative proper motion and flight direction for all the VFTS O/B-type and Wolf-Rayet stars with
\emph{Gaia} DR2 errors on the proper motion
components of less than $0.1\,\mathrm{mas\ yr^{-1}}$ (dark blue lines,
including VFTS682), and less than $0.05\,\mathrm{mas\ yr^{-1}}$ (light hatched
blue).

% Figure~\ref{fig:dist} illustrates the kinematic properties of VFTS OB-type
% and Wolf-Rayet stars in the 30 Doradus region with \emph{Gaia} DR2 errors on the proper motion
% components of less than $0.1\,\mathrm{mas\ yr^{-1}}$ (dark blue lines,
% including VFTS682), and less than $0.05\,\mathrm{mas\ yr^{-1}}$ (light hatched
% blue). The green vertical bands show the expectations for VFTS682 if
% it were ejected from R136 early in its lifetime. We highlight this star together with the two O-type
% runaways, VFTS16 and VFTS72, \mbox{(re-)identified} by \cite{lennon:18}.

% The left panel shows the distribution in proper motion relative
% to R136 from \emph{Gaia} DR2, selecting stars with small error bars only.
% Although VFTS682 is not peculiar, its relative proper motion
% falls in the range expected for it to reach its present day isolation
% within its apparent age (green vertical band). This suggests it is indeed a ``slow-runaway'' as
% suggested by \cite{bestenlehner:11}.
Subtracting
the mean motion of R136, we obtain relative proper motions (projected
velocities) of $\delta \mu_{Gaia}=0.13\pm 0.09\masyr (32\pm 21\kms)$ and
$\delta \mu_{HST}=0.20\pm 0.10\masyr (47 \pm 24\kms )$. Both values
are consistent with each other, but also with no relative motion
within $2\sigma$.

@@

% The right panel shows the distributions in projected flight direction for these
% stars: $\theta$ is the angle between the projected radial direction from the
% center of R136 to each star and its projected relative proper motion (see also \Figref{fig:main}). Dynamical ejections from the
% cluster should %are expected to
% produce
% close to radial ejections, i.e.~$\theta\simeq0$. On this panel, the green band highlights
% % the range of angles corresponding to
% an opening angle around the
% projected radial direction of 45\,degrees. The direction of motion of VFTS682 % is
% % poorly constrained because of the large errors on the proper motion
% % components both in HST and \emph{Gaia} data, but
% % it
% is consistent with dynamical ejection from R136.
We note that
VFTS72 has a small radial velocity, while VFTS16 (and possibly VFTS682) has
large peculiar radial velocity and therefore accurate distances along
the line of sight are
needed to contrain the flight direction in three dimensions.

@@

% Therefore, although the motion of VFTS682 is not an outlier compared
% to other massive stars with small proper motion errors in \emph{Gaia} DR2,
% it is possible to tentatively claim that
\todo{from here}
it is
a slow runaway with velocity compatible with dynamical ejection from
R136 early in the evolution of both the star and the cluster.

\begin{figure*}%[tbp]
  \centering
  \includegraphics[width=\textwidth]{./figures/Figure2_paper_vfts682_hst_corr_nor136err}  
  \caption{The thick yellow arrows indicate proper motions relative R136 (blue
    circle) from \emph{Gaia} DR2, multiplied by 0.4\,Myr. The prolongations in the opposite
    direction are proportional to the apparent ages of the stars, and
    the thin lines illustrate the error cone on the potential
    origin. The green thick arrow
    indicates the HST proper motion of VFTS682, with the relative
    error cone and prolongation proportional to its apparent
    age. The two blue circles indicate the regions of radii 0.01 and 0.05
    degrees around the core of R136.}

  \label{fig:main}
\end{figure*}

@@

In \citet{lennon:18} we carried out a similar study on the fast
moving O-type stars in the region, and found two very massive runaway stars in the 30 Doradus region. One of them (VFTS 16)
was previously known as a runaway star from its line of sight velocity
\citep[][]{evans:10}. \citet{lennon:18} also concluded that VFTS16 is 
the result of a dynamical ejection from the R136 cluster, while the
origin of the other star (VFTS72) is less clear given its direction
of motion (cf.~\Figref{fig:dist}). The value of $\tau_\mathrm{kin}\simeq0.9$\,Myr we find for
VFTS682 is smaller than the
corresponding value for VFTS16 of $\sim$1.5\,Myr, possibly in tension with the apparent age of that star. This means that the more
massive VFTS682 was ejected later than VFTS16 from the same
cluster. We note that also Galactic clusters can eject massive stars,
and \emph{Gaia} data have been used to study the population
surrounding Westerlund 2 \citep[][]{drew:18}.
